---
title: "In-class Exercise 01"
date: "13 January 2024"
date-modified: "last-modified"
format: html
execute: 
  echo: true # all code chunks will appear
  eval: true # all code chunks will run live
  warning: false # do not display warning message
editor: visual
---

# 1. Getting Started

First, we will download the *School questionnaire data file* from the [PISA 2022 Database](https://www.oecd.org/pisa/data/2022database/). As there is only the SAS and SPSS versions available, we will download the SAS version.

## 1.1. Loading Relevant Packages

Next, We will import the dataset into the R environment. In order to import a SAS file into R, we have to first load the `haven` package. We will do so by calling the `p_load()` function from the `pacman` package.

In addition to `haven`, we will load `tidyverse` where the `ggplot2` package can be found. We will use `haven` to load the SAS file and `ggplot2` to plot the statistical graphs.

```{r}
pacman::p_load(haven, tidyverse)
```

## 1.2. Import Data

We will now use the `read_sas` function from the `haven` package to load the dataset into R as stu_quest:

```{r}
stu_quest <- read_sas(data_file = "data/cy08msp_sch_qqq.sas7bdat")
```