---
title: "Hands-on Exercise 05 - Multivariate Analysis with Parallel Coordinates Plot"
date: "3 February 2024"
date-modified: "last-modified"
format: html
execute: 
  echo: true # all code chunks will appear
  eval: true # all code chunks will run live
  warning: false # do not display warning message
editor: visual
---


# 1. Overview

In the fourth part of the fifth lesson, we learn to visualise multivariate, numerical datau sing parallel coordinates plots. It is ideal for comparing multiple variables and seeing relationships between them. It can be used to characterise clusters that are detected during customer segmentation. 

# 2. Getting Started

## 2.1. Installing and Loading Packages

We will load the following R packages for this exercise using the `p_load()` function from the `pacman` package:

```{r}
pacman::p_load(tidyverse, GGally, parallelPlot)
```

## 2.2. Import Data

Next, we will read the data provided by the Course Instructor into the R environment. We will use the `read_csv()` function from the `readr` package found in `tidyverse`.

```{r}
wh <- read_csv("data/WHData-2018.csv")

glimpse(wh)
```

Using `glimpse` to view the data, it is observed that the data contains the happiness index in 12 columns and 156 observations.

# 3. Static Parallel Coordinates Plot

::: panel-tabset
#### Simple Parallel Coordinates Plot

Firstly, we plot a static parallel coordinates plot uisng the `ggparcoord()` function from the `GGally` package:

```{r}
ggparcoord(data = wh, 
           columns = c(7:12))
```

We are unable to find much meaning of the World Happiness measures from the plot above. 

#### With Boxplot

We can add a boxplot to the parallel plot using the arguments of the `ggparcoord()` function to show more insights:

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,     # group observations by Region column
           scale = "uniminmax", # scale variables using uniminmax where min=0 max=1
           alphaLines = 0.2,    # reduce opacity of line colour
           boxplot = TRUE,      # turn on box plot
           title = "Parallel Coordinates Plot of World Happines Variables")
```

#### With Facet

As `ggparcoord()` is an extension of the `ggplot2` package, we can include some ggplot2 functions such as `facet_wrap()` when plotting parallel coordinates plot. In this example code, each plot represents one geographical region.
```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1)) # display x-axis labels clearly
```

:::

# 4. Interactive Parallel Coordinates Plot

In this section, we create interactive parallel coordinates plots using the `parallelPlot` package. 

::: panel-tabset

#### Simple Interactive Plot

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250,
             rotateTitle = TRUE,       # rotate axis labels
             continuousCS = "Rainbow") # change colour scheme
```

#### With Histogram

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)

```
:::