---
title: "Hands-on Exercise 05 - Multivariate Analysis with Parallel Coordinates Plot"
date: "3 February 2024"
date-modified: "last-modified"
format: html
execute: 
  echo: true # all code chunks will appear
  eval: true # all code chunks will run live
  warning: false # do not display warning message
editor: visual
---


# 1. Overview

In the first part of the fourth lesson, we learn to visualise distributions using ridgeline and raincloud plots by using `ggplot2` and its extensions.

# 2. Getting Started

## 2.1. Installing and Loading Packages

We will use the following R packages in addition to `tidyverse` and `ggthemes` for this exercise:

1.  [`ggdist`](https://mjskay.github.io/ggdist): `ggplot2` extension designed for visualising distributions and uncertainty

2.  [`ggridges`](https://wilkelab.org/ggridges): to create partially overlapping line plots to visualise changes in distribution over time or space

3.  [`colorspace`](https://cran.r-project.org/web/packages/colorspace/vignettes/colorspace): toolbox for manipulating colours and palettes

The packages to be used can be installed and loaded into the R environment using the `p_load()` function from the `pacman` package:

```{r}
pacman::p_load(tidyverse, ggthemes, 
               ggdist, ggridges, colorspace)
```

## 2.2. Import Data

Next, we will read the data provided by the Course Instructor into the R environment. We will use the `read_csv()` function from the `readr` package found in `tidyverse`.

```{r}

```

Using `glimpse` to view the data, it is observed that the data contains the examination grades of a batch of primary 3 students for the subjects English, Maths, and Science.

# 3. 

::: panel-tabset
#### `geom_density_ridges()`

```{r}
#| code-fold: true
#| code-summary: "Codes here"

```

#### Varying Fill Colours


```{r}
#| code-fold: true
#| code-summary: "Codes here"


```

#### Mapping Probabilities


```{r}
#| code-fold: true
#| code-summary: "Codes here"


```

#### Mapping Quantile Lines


```{r}
#| code-fold: true
#| code-summary: "Codes here"


```

#### Mapping Specified Quantile Cut-Points



```{r}
#| code-fold: true
#| code-summary: "Codes here"

```
:::